---
title: "General information"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General information}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, warning = FALSE, message = FALSE}
library(humind)
library(dplyr)
library(tidyr)
```

This vignette emphasizes how to add indicators for the "main" dataset at the household-level. The package provides functins to add, recode, and prepare the usual humanitarian composite indicators as well as the ones that are specific to the MSNA analytical framework.

```{r load-data, echo = F, warning = FALSE, message = FALSE}
data("dummy_raw_data")
```


```{r load-data-prep}
main <- dummy_raw_data$main |> as_tibble()
loop <- dummy_raw_data$roster |> as_tibble()
```


# Recategorize demographic variables

## Head of household - final values

```{r hoh-final}
# Recategorize the head of household variable
main <- add_hoh_final(main)

# Print results
#main |> select(resp_gender, resp_age, resp_hoh_yn, hoh_gender, hoh_age) |> drop_na()

```


## Age variables

```{r age}
# Cateogize age of the respondent
main <- add_age_cat(main, "resp_age")
main <- add_age_18_cat(main, "resp_age")

# Do the same for the individuals
loop <- add_age_cat(loop, "ind_age")
loop <- add_age_18_cat(loop, "ind_age")

# Print results for the loop
#roster |> select(ind_age, ind_age_cat, ind_age_18_cat, ind_age_18_cat_d)

```
## Adding school-aged children count

```{r school-aged-children}
# Add school-aged children count to the loop
loop <- add_loop_age_dummy(loop, "ind_age", 5, 18)
main <- add_loop_age_dummy_to_main(main, loop, "ind_age_5_18")
# main |> select(ind_age_5_18_n)

# Add school_aged_female children count to the loop
loop <- add_loop_age_gender_dummy(loop, "ind_age", 5, 18, "ind_gender", "female")
main <- add_loop_age_dummy_to_main(main, loop, "ind_age_female_5_18")
# main |> select(ind_age_female_5_18)
```
