---
title: "Household-level indicators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Household-level indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(humind)
library(dplyr)
```

This vignette emphasizes how to add indicators for the "main" dataset at the household-level. The package provides functins to add, recode, and prepare the usual humanitarian composite indicators as well as the ones that are specific to the MSNA analytical framework.

```{r load-data, echo = FALSE, render=F}
data("dummy_raw_data")
```


```{r load-data-prep, render=F}
main <- dummy_raw_data$main |> as_tibble()
main$weight <- 1
```

# Shelter and NFI

## Shelter types

```{r shelter-type, render=F}
# Category of shelter type
main <- add_shelter_type_cat(main)
# main |> select(snfi_shelter_type, snfi_shelter_type_individual, snfi_shelter_type_cat)

# Shelter issues
main <- add_shelter_issue_cat(main)
# main |> select(snfi_shelter_issue, snfi_shelter_issue_n, snfi_shelter_issue_cat)
```

## Functional domestic space

```{r fds, render=F}
# Category of functional domestic space
main <- add_fds_cannot_cat(main)
# main |> select(snfi_fds_cooking, snfi_fds_sleeping, snfi_fds_storing, snfi_fds_personal_hygiene, energy_lighting_source, snfi_fds_cannot_n, snfi_fds_cannot_cat)
```


# WASH - Water, Sanitation, and Hygiene

## Sanitation facilities

```{r sanitation, render=F}
# Category of sanitation facilities
main <- add_sanitation_facility_cat(main)
# main |> select(wash_sanitation_facility, wash_sanitation_facility_cat)

# Sharing the sanitation facility
main <- add_sharing_sanitation_facility_cat(main)
# main |> select(wash_sanitation_facility_sharing_yn, wash_sharing_sanitation_facility_cat)

# Number of individuals sharing the sanitation facility
main <- add_sharing_sanitation_facility_num_ind(main)
# main |> select(wash_sanitation_facility_sharing_yn, hh_size, wash_sanitation_facility_sharing_n, wash_sharing_sanitation_facility_n_ind)
```

## Water source and time to fetch water
```{r water-source, render=F}
# Category of water source
main <- add_drinking_water_source_cat(main)

# Time to fetch water
main <- add_drinking_water_time_cat(main)

# Time to fetch water according to the 1h threshold
main <- add_drinking_water_time_threshold_cat(main)
#main |> select(wash_drinking_water_source, wash_drinking_water_time_yn, wash_drinking_water_time_int, wash_drinking_water_time_sl, wash_drinking_water_time_cat, wash_drinking_water_time_threshold_cat)
```

