---
title: "Example of survey analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example of survey analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installattion requirements

If `impactR.analysis`, `impactR.kobo`, and `impactR.viz` are not installed, you can install it like so:

```{r}
# pak::pak("gnoblet/impactR.analysis")
# pak::pak("gnoblet/impactR.kobo")
# pak::pak("impact-initiatives-hppu/impactR.viz")
```

# Load libraries an files

Load packages:
```{r setup, warning = FALSE, message = FALSE}
library(humind)
library(impactR.kobo)
library(impactR.analysis)
library(dplyr)
library(impactR.viz)
```

Load and prepare data, and add indicators:
```{r prep-data}
main <- humind::dummy_raw_data$main
main$weight <- 1

# Add all indicators needed (eventually this will be done in a one line pipe)
main <- main |> 
  # Demographics
  add_hoh_final() |> 
  add_age_cat("resp_age") |> 
  add_age_18_cat("resp_age") |> 
  add_age_cat("hoh_age") |> 
  add_age_18_cat("hoh_age") |>
  # WASH
  add_sanitation_facility_cat() |> 
  add_sharing_sanitation_facility_cat() |> 
  add_sharing_sanitation_facility_num_ind() |> 
  add_drinking_water_source_cat() |> 
  add_drinking_water_time_cat() |>
  add_drinking_water_time_threshold_cat() |>
  # SNFI
  add_shelter_type_cat() |>
  add_shelter_issue_cat() |> 
  add_fds_cannot_cat() |> 
  add_occupancy_cat()

# Make it a survey design (IRL please use your survey design)
main <- srvyr::as_survey_design(main, weights = "weight", stratum = "admin1")
```

Now we load the kobo tool and data analysis plan:
```{r prep-analysis, warning=F}
survey <- humind::survey_update |> 
  rename(label = label_english) |> 
  split_survey(type)
choices <- humind::choices_update |> 
  rename(label = label_english)
analysis_dap <- humind::analysis_dap
```


# Analysis

```{r run-analysis, warning=F}
# Run the analysis
an <- impactR.analysis::kobo_analysis_from_dap(main, analysis_dap |> filter(dataset == "main"), survey, choices)
# If we want a grouped analysis by admin 1
an_ad1 <- impactR.analysis::kobo_analysis_from_dap(main, analysis_dap |> filter(dataset == "main"), survey, choices, group = "admin1")
```

# Visualization of results

## Simple bar graphs

Let's do one quick viz:
```{r viz-bar, warning = FALSE, out.width = "65%", fig.align = "center"}
# Drinking water sources for the whole country
var_to_plot <- "wash_drinking_water_source_cat"

# Filter the analysis results
an_plot <- an |> 
  filter(var == var_to_plot)

# Get title, total number of observations and number of missing values
title_to_plot <- an_plot$indicator |> unique()
n_tot_plot <- an_plot$n_tot |> unique()
na_count_to_plot <- an_plot$na_count_tot |> unique()

# Bar plot
an_plot |> 
  mutate(
    stat_r = round(stat*100, 1),
    var_value_na = factor(ifelse(is.na(var_value), "missing", var_value), levels = c("missing", "undefined", "improved", "unimproved", "surface_water"))
    ) |> 
  bar(
    x = "var_value_na",
    y = "stat_r",
    add_text = F, 
    title = title_to_plot, 
    theme_fun = theme_reach(title_size = 12),
    caption = paste0("Number of observations: ", n_tot_plot, " (", na_count_to_plot, " missing)"))

```

Let's do one quick viz for the grouped analysis at the admin 1 level:
```{r viz-bar-grouped, warning = FALSE, out.width = "65%", fig.align = "center"}
# Drinking water sources for the whole country
var_to_plot <- "wash_drinking_water_source_cat"

# Filter the analysis results
an_ad1_plot <- an_ad1 |> 
  filter(var == var_to_plot, group_key == "admin1")

# Get title, total number of observations and number of missing values
title_to_plot <- an_plot$indicator |> unique()
n_tot_plot <- an_plot$n_tot |> unique()
na_count_to_plot <- an_plot$na_count_tot |> unique()

# Stack bar plot
an_ad1_plot |> 
  mutate(
    stat_r = round(stat*100, 1),
    var_value_na = factor(ifelse(is.na(var_value), "missing", var_value), levels = c("missing", "undefined", "improved", "unimproved", "surface_water"))
    ) |> 
  bar(
    x = "admin1", 
    y = "stat_r", 
    group = "var_value_na", 
    position = "stack", 
    add_text = T, 
    add_text_color = "white",
    add_text_threshold_display = 5,
    title = title_to_plot, 
    theme_fun = theme_reach(
      grid_major_x = FALSE,
      axis_x = FALSE, 
      axis_ticks_y = FALSE,
      legend_direction = "horizontal", 
      legend_position = "bottom", 
      title_size = 14),
    caption = paste0("Number of observations: ",
                     n_tot_plot, " (", na_count_to_plot, " missing)"))
```
